---
output:
  pdf_document:
    includes:
      in_header: preamble.tex

  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

source("1-load-data.R")
source("2-preparation.R")
source("3-modeling.R")

library(RColorBrewer)

```

# Policy brief: Europeans and False Online Information

\textit{Document created `r Sys.Date()`}

## Strengthening Digital Fact-Checking Skills in the Age of AI

The rise of false, misleading, and deliberately deceptive digital content, often amplified by artificial intelligence, poses a growing threat to liberal democracy, human rights, and core European values.

\begin{center}
\begin{tcolorbox}[orangebox]
This brief aims to answer the following questions:

\begin{enumerate}
  \item What influences Europeans’ \textbf{confidence in digital fact-checking}, i.e., verifying the truthfulness of online information and content?
  \item What \textbf{policy recommendations can be formulated to strengthen the digital fact-checking skills} of European citizens?
\end{enumerate}
\end{tcolorbox}
\end{center}

## The Confidence Gap

\begin{flushleft}
The CRONOS 3 survey asked people across Europe to what extent the statement: 
“\textbf{I know how to check the truthfulness of the information or content I find on the internet}”
is true for them. \textcolor{exactgreen}{\textbf{Only 30\% said it was very true, thus they were confident in digital fact-checking skills}}. 
At the same time, as many as 86\% say they use the Internet almost all the time or several times a day. 
That means \textcolor{exactorange}{\textbf{67\% of most frequent internet users are not confident in their digital fact-checking skills}}.
\end{flushleft}

```{r xtab, include=FALSE}
library(webshot)

c_mutated_tab <- c_mutated %>% 
  filter(no_int_access_or_use == 0) %>% 
  select(w1weight, TRC, w1dq3) %>% 
  mutate(NETU_T = case_when(
    w1dq3 == 1 | w1dq3 == 2 ~ 1, 
    is.na(w1dq3) ~ NA_integer_, 
    .default = 0), 
    NETU_T = set_label(NETU_T, "Uses internet almost all the time or several times a day"), 
    TRC = set_label(TRC, "Knows how to check truthfulness of online content"), 
    NETU_T = set_labels(NETU_T, labels = c("Yes" = 1, "No" = 0)), 
    TRC = set_labels(TRC, labels = c("Very true" = 1, "Other" = 0)))  
  

tab_xtab(c_mutated_tab$TRC, c_mutated_tab$NETU_T,
         weight.by = c_mutated_tab$w1weight, 
         show.na = T, 
         show.obs = F, 
         show.col.prc = T, 
         show.row.prc = T, 
         show.summary = F,
         emph.total = T, 
         use.viewer = F,
         digits = 0,
         file = "t.html")

webshot("t.html", "t.pdf", vheight = 200, vwidth = 400) 
```

## Factors of Confidence in Digital Fact-Checking

We analyzed several factors that influence respondents’ confidence in digital fact-checking. We found meaningful associations between other digital skills and two types of trust:

```{r logit-plot, fig.height=3.5, message=FALSE}

non_sig_M2 <- c("PGW", "SMP", "NETUALL", "DIGSKILLT", "GDF_C", "EDY_C", "AGE_C", "HTI_C")


lbl <- c(
  IFS = "Find product or service \ninformation online",
  LMA = "Limit access \nto social media profile",
  BSC = "Restrict browser \ncookie settings",
  FLC = "Create digital \nfiles",
  SCNTR = "Trust in scientists",
  GENTR = "General trust \nin others"
)


p1 <- plot_model(
  M2,
  show.values = T,
  value.offset = 0.4,
  colors = "Set2",
  rm.terms = non_sig_M2,
  sort.est = T,
  axis.labels = "",
  axis.lim = c(0.6, 5), 
  title = ""
) +
  labs(
subtitle = "People confident in finding information about goods or services online,\nlimiting access to their social media profiles, and restricting cookies\nare more likely to be confident in digital fact-checking.\nTrust in scientists also increases this confidence; general trust reduces it.", 
caption = "Source: Own elaboration based on CRONOS3 and ESS10–11 survey data.\nDots represent odds ratio values; lines indicate the uncertainty of the results.") + 
  theme_minimal() + 
  geom_hline(yintercept = 1, linetype = "dashed", color = "darkgray", alpha = 0.6) +
  annotate("text", x = 3.8, y = 0.85,
           label = "Odds Ratio of 1 = no effect", color = "darkgray", alpha = 0.6) +
  annotate("text", x = 0.7, y = 1,
           label = "Below 1 = negative effect", color = "#66C2A5" , alpha = 0.6) + 
  annotate("text", x = 4.5, y = 6,
           label = "Above 1 = positive effect", color = "#FC8D62" , alpha = 0.6) + 
  scale_x_discrete(labels = lbl)

plot(p1)


```

~This work is licensed under CC BY 4.0. To view a copy of this license, visit [creativecommons.org/licenses/by/4.0](https://creativecommons.org/licenses/by/4.0/)~

\begin{center}
\includegraphics[width=44mm]{by.png}
\end{center}

## Digital Skills Matter

The most important factor is confidence in finding information about goods or services online. People who declare this belief have **over 3.5 times the odds** of being confident in fact-checking **than those who do not**. This suggests that people look at information truthfulness as consumers.

\begin{center}
\begin{tcolorbox}[greenbox]
Recommendation 1:

\textbf{Design fact-checking campaigns and training that mirror familiar online shopping behaviors. This makes the skill feel more intuitive and easier to learn}.
\end{tcolorbox}
\end{center}

The next essential factors are limiting access to one’s profile or content on social media and restricting cookies via internet browsers. These digital privacy skills increase confidence by **2.9 and 2.3 times the odds**.

\begin{center}
\begin{tcolorbox}[orangebox]
Recommendation 2:

\textbf{Combine fact-checking campaigns and training with digital privacy issues. These skills reinforce each other and may have similar motivations}.
\end{tcolorbox}
\end{center}

We found that participation in any digital skills training does not affect confidence in digital fact-checking. Thus, the training needs to be focused on that skill set. However, basic digital know-how does matter. People confident in creating digital files have **almost 2 times the odds** of being confident in fact-checking.

## The Surprising Role of Trust

**Each step up of general trust in others decreases the likelihood** of confidence in digital fact-checking **by 11%** (0.89 odds ratio). The more trusting the respondents are, the less likely they are to consider themselves capable of fact-checking online.

\begin{center}
\begin{tcolorbox}[greenbox]
Recommendation 3:

\textbf{Highlight the difference between trusting people in real life and trusting sources online.\\The internet is neither just an extension of the offline world nor has the same social rules}.
\end{tcolorbox}
\end{center}

**Each step up of trust in scientists** **increases the likelihood** of confidence in digital fact-checking **by 9%** (1.09 odds ratio). Therefore, fact-checking skills and trust in people who provide evidence-based and validated information are positively related.

\begin{center}
\begin{tcolorbox}[orangebox]
Recommendation 4:

\textbf{Invest in academia to support scholars in sharing their work in accessible ways.\\Train academic researchers to communicate their findings to the public}.
\end{tcolorbox}
\end{center}

### Analytical note

\begin{flushleft}
\footnotesize

We studied 14 factors of confidence in digital fact-checking, and 8 had no meaningful impact: computer programming skills, using social media, frequency of internet use, participation in digital skills training, gender, age, education, income, and country of residence.

Open data from CRONOS3 Wave 1 (\url{https://doi.org/10.21338/cronos3-w1}), ESS10 (\url{https://doi.org/10.21338/ess10e03_2}), ESS10sc (\url{https://doi.org/10.21338/ess10sce03_1}), and ESS11 (\url{https://doi.org/10.21338/ess11e03_0}) represent countries: Austria, Belgium, Czechia, Finland, France, Hungary, Iceland, Poland, Portugal, Slovenia, and the United Kingdom. Respondents who did not use the internet or had no access to it were excluded; a total unweighted sample in the analysis was $n = 9{,}538$.

Supplementary tables and code to reproduce our analysis are openly available at: \url{https://github.com/zremek/I4NG-Hackathon-2025}
\normalsize
\end{flushleft}

~**Authors**: Jakub Adamski, [jakub.adamski\@uni.lodz.pl](mailto:jakub.adamski@uni.lodz.pl) Remigiusz Żulicki, [remigiusz.zulicki\@uni.lodz.pl](mailto:remigiusz.zulicki@uni.lodz.pl)~

~[Faculty of Economics and Sociology](https://www.eksoc.uni.lodz.pl/en/about-the-faculty/get-to-know-eksoc), University of Lodz, Poland~
